Bees need larger brains to thrive in urban environments

### 2023-09-06 Manuscript/Supplementary_material.Rmd
Response to R

---
output: pdf_document
header-includes:
- \usepackage{orcidlink}
- \usepackage{caption,setspace}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{float}
- \floatplacement{figure}{H}
---

@@

# Supplementary material

## Brain and body size measurements

@@

@@

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.align="center", fig.height=12, fig.width=12, fig.cap="\\textbf{Figure S2.} Maps showing the study areas and the number of GBIF occurrences in logarithmic scale for the selected bee species. A) East Coast of United States with records for Vermont, New Hampshire, Massachusetts, Rhode Island, Connecticut, New York metropolitan area, New Jersey, Delaware and Maryland. B) European continent with records downloaded for England, Belgium, Netherlands, Germany and Denmark.",  out.width="90%"}

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,fig.align="center", fig.height=5, fig.width=12, fig.cap="\\textbf{Figure S3.} Frequency distribution of the bee preferences for each habitat type (i.e., natural, agricultural and urban). The different habitats are reclassifications from the 2006 land use categories of the Corine Land Cover (CLC) inventory for Europe and the National Land Cover (NLC) database for Unites States.", fig.pos="p"}

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,fig.align="center", fig.height=6, fig.width=8, fig.cap="\\textbf{Figure S4}. Association between relative brain size (A and D), brain weight (B and E) and intertegular distance (C and F) with habitat preference separated by habitat type (i.e., natural, agricultural and urban) and geographical region (USA in the upper panel and Europe in the lower one). The shaded and coloured areas by habitat type represent 95\\% credible intervals.", fig.pos="p", out.width="100%"}

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,fig.align="center", fig.height=6, fig.width=8, fig.cap="\\textbf{Figure SX}. ", fig.pos="p", out.width="100%"}

@@

#Natural
p1_1 = ggplot(d, aes(x = Lecticity, y = Natural)) + 
geom_violin(alpha=0.4, fill="#d7e1ee",linewidth=0)+
geom_boxplot(colour="black",fill="grey",
width = .15, 
outlier.shape = NA) +
geom_jitter(alpha=0.3,width = 0.1) +
ylab("Habitat Preference")+
xlab(NULL)+
ggtitle("Natural")+
theme_bw()

#Agricultural
p1_2 = ggplot(d, aes(x = Lecticity, y = Agricultural)) + 
geom_violin(alpha=0.4, fill="#d7e1ee",linewidth=0)+
geom_boxplot(colour="black",fill="grey",
width = .15, 
outlier.shape = NA) +
geom_jitter(alpha=0.3,width = 0.1) +
ylab(NULL)+
xlab("Diet")+
ggtitle("Agricultural")+
theme_bw()

#Urban
p1_3 = ggplot(d, aes(x = Lecticity, y = Urban)) +
geom_violin(alpha=0.4, fill="#d7e1ee",linewidth=0)+
geom_boxplot(colour="black",fill="grey",
width = .15, 
outlier.shape = NA) +
geom_jitter(alpha=0.3,width = 0.1) +
ylab(NULL)+
xlab(NULL)+
ggtitle("Urban")+
theme_bw()

@@

p2_1 = ggplot(d1, aes(x = Lecticity, y = residuals)) + 
geom_violin(alpha=0.4, fill="#d7e1ee",linewidth=0)+
geom_boxplot(colour="black",fill="grey",
width = .15, 
outlier.shape = NA) +
geom_jitter(alpha=0.3,width = 0.1) +
ylab("Relative brain size")+
xlab("Diet")+
theme_bw()

p2_2 = ggplot(d1, aes(x = Lecticity, y = Brain.weight)) + 
geom_violin(alpha=0.4, fill="#d7e1ee",linewidth=0)+
geom_boxplot(colour="black",fill="grey",
width = .15, 
outlier.shape = NA) +
geom_jitter(alpha=0.3,width = 0.1) +
ylab("Absolute brain size")+
xlab("Diet")+
theme_bw()

panel2 =  p2_2 + p2_1

@@

plot_grid(plot_grid(panel1, labels = c('A'))+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1)),

plot_grid(panel2, labels = c('B'))+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1)), nrow = 2)

@@

>>>>

---
output: pdf_document
header-includes:
- \usepackage{orcidlink}
- \usepackage{caption,setspace}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{float}
- \floatplacement{figure}{H}

---

@@

# Supplementary material for: Bees need larger brains to thrive in urban environments

## Authors: Jose B. Lanuza, Miguel Á. Collado, Ferran Sayol, Daniel Sol and Ignasi Bartomeus

## Supplementary text S1: Brain and body size measurements

@@

**References**:

@@

## Supplementary text S2: Intra- and inter-specific brain body size variation

In average we had 3.31 individuals measured per species with information for both brain and body size. These different individuals within a species tended to have similar values of brain and body size (i.e., low variance) in comparison with the variability found across species. To show this pattern, we calculated the Intraclass Correlation Coefficient (ICC) that estimates how similar or different are the values within the different groups (i.e., species). The ICC values range from 0 to 1. For our case, values close to 0 indicate that the different individual measurements per species tend to be very different to each other and values close to 1 indicate that the measurements within a species tend to be very similar or identical. In order to compute the ICC, we first fitted an intercept-only linear model with the help of the *lmer* function from the *lme4* library version 1.10-34 (Bates, 2010) where we only included a response variable of interest (i.e., brain or body size) and species as a random effect. This allowed us to calculate the variances of interests in order to compute the ICC. We calculated the ICC with the help of the *icc* function from the library *performance* version 0.10.3 (Lüdecke et al., 2021). This function computes the division between the random effect variance and the total variance (i.e., the sum of the random effect variance and the residual variance). We found that for both brain and body size the individual measurements per species tended to be very similar between each other (ICC = 0.947; ICC ICC = 0.931, respectively) but we found high variability across species (see Figure S4). Overall, this indicates that the mean values of brain and body size have biological significance and provides support for investigating macro-ecological patterns across species.

**References**:

Bates, D. M. (2010). lme4: Mixed-effects modeling with R.

Lüdecke, D., Ben-Shachar, M. S., Patil, I., Waggoner, P., & Makowski, D. (2021). performance: An R package for assessment, comparison and testing of statistical models. Journal of Open Source Software, 6(60).

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,fig.align="center", fig.height=6, fig.width=8, fig.cap="\\textbf{Figure S2}. Graphical representation and statistical differences of diet specialisation (i.e., oligolectic and polylectic species) by habitat preference per habitat type (Panel A) and brain size including absolute and relative brain size (Panel B). The graphical representation shows boxplots, raw values and a shaded violin plot to allow a better understanding of the spread, density and number of values by each category. Note that the raw values are jittered horizontally in order to allow a better visualisation of overlapping points. The significance of the statistical differences across values are shown with the resulting P-values of a Wilcoxon test. Interestingly, we see that (i) there are few oligolectic species represented in our dataset; (ii) there are no statistical differences between oligolectic and polylectic species in habitat preference (Panel A) and brain size (Panel B); and, (iii) oligolectic species tend to have larger relative brains and, if anything, show lower preferences by urban habitats, indicating that feeding specialisation is unlikely to be driving the observed association between urban preference and brain size.", fig.pos="p", out.width="100%"}

@@

#Plotting----
#Here we load some libraries to include Wilcoxon rank test in the plot
#for publication
library(ggpubr)
library(rstatix)

#First graph Natural habitat
stat.test1 <- d %>%
wilcox_test(Natural ~ Lecticity) %>%
add_significance()
stat.test1 <- stat.test1 %>% add_xy_position(x = "Lecticity")
#Build plot
p1_1 = ggplot(d, aes(x = Lecticity, y = Natural)) + 
geom_violin(alpha=0.4, fill="#d7e1ee",linewidth=0)+
geom_boxplot(colour="black",fill="grey",
width = .15, 
outlier.shape = NA) +
geom_jitter(alpha=0.3,width = 0.1) +
ylab("Habitat preference")+
xlab(NULL)+
ggtitle("Natural")+
stat_pvalue_manual(stat.test1, label =  "Wilcoxon test, p = {p}",vjust = -0.05, bracket.nudge.y = 0.05) +
scale_y_continuous(expand = expansion(mult = c(0.05, 0.2))) +
theme_bw()

#Second graph Agricultural habitat
stat.test2 <- d %>%
wilcox_test(Agricultural ~ Lecticity) %>%
add_significance()
stat.test2 <- stat.test2 %>% add_xy_position(x = "Lecticity")
#Build plot
p1_2 = ggplot(d, aes(x = Lecticity, y = Agricultural)) + 
geom_violin(alpha=0.4, fill="#d7e1ee",linewidth=0)+
geom_boxplot(colour="black",fill="grey",
width = .15, 
outlier.shape = NA) +
geom_jitter(alpha=0.3,width = 0.1) +
ylab(NULL)+
xlab("Diet")+
ggtitle("Agricultural")+
stat_pvalue_manual(stat.test2, label =  "Wilcoxon test, p = {p}",vjust = -0.05, bracket.nudge.y = 0.05) +
scale_y_continuous(expand = expansion(mult = c(0.05, 0.2))) +
theme_bw()

#Third graph Urban habitat
stat.test3 <- d %>%
wilcox_test(Urban ~ Lecticity) %>%
add_significance()
stat.test3 <- stat.test3 %>% add_xy_position(x = "Lecticity")
#Build plot
p1_3 = ggplot(d, aes(x = Lecticity, y = Urban)) +
geom_violin(alpha=0.4, fill="#d7e1ee",linewidth=0)+
geom_boxplot(colour="black",fill="grey",
width = .15, 
outlier.shape = NA) +
geom_jitter(alpha=0.3,width = 0.1) +
ylab(NULL)+
xlab(NULL)+
ggtitle("Urban")+
stat_pvalue_manual(stat.test3, label =  "Wilcoxon test, p = {p}",vjust = -0.05, bracket.nudge.y = 0.05) +
scale_y_continuous(expand = expansion(mult = c(0.05, 0.2))) +
theme_bw()

@@

#Plotting----
#First graph brain weight
stat.test4 <- d1 %>%
wilcox_test(Brain.weight ~ Lecticity) %>%
add_significance()
stat.test4 <- stat.test4 %>% add_xy_position(x = "Lecticity")
#Build plot
p2_1 = ggplot(d1, aes(x = Lecticity, y = Brain.weight)) + 
geom_violin(alpha=0.4, fill="#d7e1ee",linewidth=0)+
geom_boxplot(colour="black",fill="grey",
width = .15, 
outlier.shape = NA) +
geom_jitter(alpha=0.3,width = 0.1) +
ylab("Absolute brain size")+
xlab("Diet")+
stat_pvalue_manual(stat.test4, label =  "Wilcoxon test, p = {p}",vjust = -0.05, bracket.nudge.y = 0.05) +
scale_y_continuous(expand = expansion(mult = c(0.05, 0.2))) +
theme_bw()

#Second graph residuals
stat.test5 <- d1 %>%
wilcox_test(residuals ~ Lecticity) %>%
add_significance()
stat.test5 <- stat.test5 %>% add_xy_position(x = "Lecticity")
#Build plot
p2_2 = ggplot(d1, aes(x = Lecticity, y = residuals)) + 
geom_violin(alpha=0.4, fill="#d7e1ee",linewidth=0)+
geom_boxplot(colour="black",fill="grey",
width = .15, 
outlier.shape = NA) +
geom_jitter(alpha=0.3,width = 0.1) +
ylab("Relative brain size")+
xlab("Diet")+
stat_pvalue_manual(stat.test5, label =  "Wilcoxon test, p = {p}",vjust = -0.05, bracket.nudge.y = 0.05) +
scale_y_continuous(expand = expansion(mult = c(0.05, 0.2))) +
theme_bw()

panel2 =  p2_1 + p2_2

@@

plot_grid(plot_grid(panel1, labels = c('A'))+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=2)),

plot_grid(panel2, labels = c('B'))+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=2)), nrow = 2)

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.align="center", fig.height=12, fig.width=12, fig.cap="\\textbf{Figure S3.} Maps showing the study areas and the number of GBIF occurrences in logarithmic scale for the selected bee species. A) East Coast of United States with records for Vermont, New Hampshire, Massachusetts, Rhode Island, Connecticut, New York metropolitan area, New Jersey, Delaware and Maryland. B) European continent with records downloaded for England, Belgium, Netherlands, Germany and Denmark.",  out.width="90%"}

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,fig.align="center", fig.height=6, fig.width=8, fig.cap="\\textbf{Figure S4}. Mean and standard devaition of brain weight and intertegular distance for all the species included in our analyses (N=89).", fig.pos="p", out.width="100%"}

#Calculate intraclass correlation coefficiet
#for brain and body size
#Exploration of the dataset (number of replicates)
#and show graphically interspecific variation

#Load libraries
library(dplyr)
library(stringr)
library(readr)
library(lme4)
library(performance)

#Read data
long = read_csv("../Raw_data/all_cleaning.csv")

#Filter out the species that we don't use
#Load preferences data
preferences_usa = readr::read_csv("../Data/USA_data/preferences_usa.csv") 
preferences_eu = readr::read_csv("../Data/Europe_data/preferences_europe.csv") 
#Bind rows
pref = bind_rows(preferences_usa, preferences_eu)
spp = unique(pref$Species)

#Select just the species that we used in the study for calculating preferences
#after applying our filtering criteria (see mthods for further justification)
long = long %>% 
filter(Species %in% spp)

#Show graphically variation in the distribution across species 
#For both brain and body sizes
#1st for brain weight
d = long %>% 
group_by(Species) %>% 
summarise(mean = mean(Brain.weight), std = sd(Brain.weight)) %>% 
arrange(desc(mean))
#Plotting
p1 = ggplot(d, aes(x=reorder(Species, mean), y=mean, fill=std)) + 
geom_point()+
geom_errorbar(aes(ymin=mean-std, ymax=mean+std), width=.2,
position=position_dodge(0.05)) +
theme_bw() +
coord_flip() +
theme(legend.position = "none",axis.text.y = element_text(size = 4),
plot.title = element_text(face = "bold")) +
xlab("Species") +
ylab("Brain weight (mg)") +
ggtitle("A")

#2nd for body size
d1 = long %>% 
group_by(Species) %>% 
summarise(mean = mean(IT), std = sd(IT)) %>% 
arrange(desc(mean))
#Plotting
p2 = ggplot(d1, aes(x=reorder(Species, mean), y=mean, fill=std)) + 
geom_point()+
geom_errorbar(aes(ymin=mean-std, ymax=mean+std), width=.2,
position=position_dodge(0.05)) +
theme_bw() +
coord_flip() +
theme(legend.position = "none",axis.text.y = element_text(size = 4),
plot.title = element_text(face = "bold")) +
xlab("Species") +
ylab("Intertegular distance (mm)")+
ggtitle("B")

p1 + p2

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,fig.align="center", fig.height=5, fig.width=12, fig.cap="\\textbf{Figure S5.} Frequency distribution of the bee preferences for each habitat type (i.e., natural, agricultural and urban). The different habitats are reclassifications from the 2006 land use categories of the Corine Land Cover (CLC) inventory for Europe and the National Land Cover (NLC) database for Unites States.", fig.pos="p"}

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,fig.align="center", fig.height=6, fig.width=8, fig.cap="\\textbf{Figure S6}. Association between relative brain size (A and D), brain weight (B and E) and intertegular distance (C and F) with habitat preference separated by habitat type (i.e., natural, agricultural and urban) and geographical region (USA in the upper panel and Europe in the lower one). The shaded and coloured areas by habitat type represent 95\\% credible intervals.", fig.pos="p", out.width="100%"}
